Changes in Patterns of Geometric Sign Groupings as a Model of Social Dynamics Across Aurignacian Phases

### 2025-03-14 endofquarterpaper.qmd

Our data comes from SignBase 1.0, an open-access collection of geometric signs on mobile objects from Upper Paleolithic Europe. We used the identification and categorization of the sign types provided by SignBase's curators [@dutkiewicz2020]. The data set consists of `r nartifacts` objects found at `r nsites` sites in `r ncountries` countries, and records `r nsigntypes` geometric sign types. We excluded signs if they fell under the "other" category, due to the wide range of potential possibilities. We also excluded seven sites from the sample (only one sign present: Willendorf; extreme geographic distance from the other sites: Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave; dated much later than the rest of the assemblages: Grotte De La Princesse Pauline and Šandalja II). These exclusions resulted in a sample of `r nartifacts_clean` objects found at `r nsites_clean` sites in `r ncountries_clean` countries, and `r nsigntypes_clean` sign types used in this study (@fig-site-map).

```{r fig.height= 9, fig.width=10}
#| label: fig-site-map
#| fig-cap: "Map of sites containing geometric signs on mobile objects from Upper Paleolithic Europe, sampled from the SignBase dataset, divided by Aurignacian phases"

@@

plot_grid(sign_map, signcount_barplot, ncol=1, align = "v", scale = c(1, 0.8))

@@

Ages of the objects were determined by calibrating the uncalibrated ages provided by SignBase, using the rcarbon package [@Crema_Bevan_2021]. The calibrated radiocarbon dates of the sampled data range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years. We assigned each artifact to a chronological phase following the date ranges recorded in @tbl-aur-table. The Transitional phase has `r trans_site_count` sites and `r trans_sign_count` signs, the Proto-Aurignacian has `r proto_site_count` sites and `r proto_sign_count` signs, the Early Aurignacian has `r early_site_count` sites and `r early_sign_count` signs, and the Evolved Aurignacian has `r evolved_site_count` sites and `r evolved_sign_count` signs.

SignBase 1.0 is a record of objects noting only the presence or absence of each sign, without recording the quantity or sequence of individual signs on a given object. The distribution of objects across sites is highly skewed, with Vogelherd having `r nvogel`, but the median number of objects is `r artifact_count_median`. To minimize the influence of this skewed distribution, we used presence/absence data for each sign at each site, reducing the effect of variation in object count between sites. To make sure that the exclusion of abundance would not significantly impact our results, we ran a Mantel tests both between a Euclidean object abundance distance matrix and a Jaccard distance matrix of the sign types [@Lycett_2019] (p = `r p_value_object_abundance_correlation`, Mantel statistic R = `r r_value_object_abundance_correlation`), and a Eudclidean sign abundance distance matrix and a Jaccard sign type distance matrix (p = `r p_value_sign_abundance_correlation`, Mantel statistic R = `r r_value_sign_abundance_correlation`) and found that neither were statistically significant. Another Mantel test was run between a Euclidean distance matrix for dates and the Jaccard sign type matrix in order to determine whether there was a definitive correlation between distance in time and distribution of sign types, finding a statistically significant correlation (p = `r chron_p`, Mantel statistic R = `r chron_r`). This confirms that a meaningful change in sign types occurs over time.

{run permanova test for chronology}

@@

Seriation analysis is the arrangement of data into a linear order to reveal patterns [@Hahsler_Hornik_Buchta]. We used the Browner-Kile seriation algorithm [@Brower_Kile_1988] from the seriation package [@seriation_pckge], in which a unidimensional sequence is generated by reordering the rows and columns to group the presences along the diagonal. Network analysis represents and quantifies the relationship between nodes [@Mills_2017], in this case, sites. In our network analysis, each node is a site, and each edge is a measure of sign type distribution similarity. We input the Jaccard similarity matrix for each phase, computed the network statistics, and plotted the results with the Fruchterman-Reingold layout [@Fruchterman_Reingold_1991], a force-directed graph layout for uniform edge lengths. Network analysis was performed using the vegan [@Vegan2001], statnet [@statnet_package] and igraph [@igraph_package] packages.

@@

To evaluate the strength of the groupings found in the seriation and network analyses, we used modularity and perMANOVA tests. The modularity test is a community structure network detection method, and quantifies the randomness of the groupings by measuring the fraction of the edges in the network that fall within groups, minus the expected value if the edges were placed at random [@Newman_2006; @Newman_Girvan_2004]. The results are expressed in a score of 0 to 1, with 0 meaning no significant subgroups, and 1 meaning perfect groupings. PerMANOVA quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017]. The results are expressed by an R-statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, and a p-value, which quantifies whether the groups are significantly different (we use an alpha value of 0.05 for the threshold of statistical significance). The modularity test and perMANOVA were performed using the vegan [@Vegan2001] and igraph [@igraph_package] packages.

@@

The isolation-by-distance framework proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019]. If cultural difference can be explained solely by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases. To test this, we determined the correlation between sign distribution and geography using Mantel tests, via the vegan package [@Vegan2001]. The Mantel tests were run with a geographic distance matrix and a Jaccard similarity matrix. The test seeks to determine the level of correspondence between the matrices through permutational evaluation of the null distribution [@Mantel_Smouse], (here 1000 permutations), producing a Mantel R statistic and a p-value. The R statistic falls between -1, meaning a strong negative correlation, and 1, meaning a strong positive correlation, while a value of 0 means no correlation. The p-value quantifies statistical significance of the correlation between the two matrices, with p = 0.05 serving as the threshold for statistical significance.

@@

For each of the phases, seriation identifies at least two groups, one with a minimal range of signs, and one with a broader range. We also see that the notch and line signs are the most prevalent signs across all time periods. In the Transitional phase, we see some variety within groups, with the first consisting of Fumane, Pod Hradem, and Geissenklösterle, two of which only have the "notch" sign, and one of which has the notch and circumnotch, and the other consisting of Labeko Koba, Hohle Fels, and El Castillo, which have a broader diversity of signs. In the Proto-Aurignacian, we see an expansion in both sites and range of signs, as well as movement to new sites, with little overlap in site names. The first group consists of Abri Pataud, Gatzarria, Hohlenstein-Stadel, and Mladeč, three of which only have the notch and line signs, and one which also has a dashline sign, and the other consists of Geissenklösterle, Spy, Vogelherd, Grotte du Renne, La Ferrassie, and Les Cottés. In the second group, we see fairly consistent patterning of a diverse range of signs, although Les Cottés varies a little. In the Early Aurignacian, we again see movement to new sites with no overlap from the Proto-Aurignacian, as well as increased diversity of patterns within the groups. The first group consists of Grottes de Fonds-de-Forêt (radnotch and notch), Riparo Bombrini (notch), and Grotte de la Verpillière I (notch and line), The second consists of Solutré, Hohle Fels, Castanet, Cellier, Blanchard, and Trou al'Wesse, which contain a wide range of signs in a variety of combinations. Lastly, in the Evolved Aurignacian, we see slightly more consistent groups, again with movement to new sites with no overlap from the Early Aurignacian. The first group consists of Les Rois, Gargas, La Viña, Sirgenstein Cave, and Vindija Cave, covering the line, notch, and and obline signs. The second consists of Trou Magrite, Bockstein-Törle, Vogelherd, and Hohle Fels. Interestingly, this is the only phase to have more sites in the group with a minimal range of signs, than in the group with a more diverse range.

>>>>

Our data comes from SignBase 1.0, an open-access collection of geometric signs on mobile objects from Upper Paleolithic Europe. SignBase 1.0 is a record of objects noting only the presence or absence of each sign, without recording the quantity or sequence of individual signs on a given object. We used the identification and categorization of the sign types provided by SignBase's curators [@dutkiewicz2020]. The data set consists of `r nartifacts` objects found at `r nsites` sites in `r ncountries` countries, and records `r nsigntypes` geometric sign types. We excluded signs if they fell under the "other" category, due to the wide range of potential possibilities. We also excluded seven sites from the sample (only one sign present: Willendorf; extreme geographic distance from the other sites: Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave; dated much later than the rest of the assemblages: Grotte De La Princesse Pauline and Šandalja II). These exclusions resulted in a sample of `r nartifacts_clean` objects found at `r nsites_clean` sites in `r ncountries_clean` countries, and `r nsigntypes_clean` sign types used in this study (@fig-site-map).

```{r fig.height= 9, fig.width=10}
#| label: fig-site-map
#| fig-cap: "A: Map of sites containing geometric signs on mobile objects from Upper Paleolithic Europe, sampled from the SignBase dataset, divided by Aurignacian phases. B: "

@@

plot_grid(sign_map, signcount_barplot, ncol=1, align = "hv", scale = c(1, 0.8), labels = "AUTO")

@@

The distribution of objects across sites is highly skewed, with the median number of objects as `r artifact_count_median`, but Vogelherd has `r nvogel`. To minimize the influence of this skewed distribution, we followed @Lycett_2019 in using presence/absence data for each sign at each site, reducing the effect of variation in object count between sites. A Mantel test between a Euclidean object abundance distance matrix and a Jaccard distance matrix of sign types indicated that there was no significant correlation between object abundance and sign type distribution (Mantel statistic R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`).

Ages of the objects were determined by calibrating the uncalibrated ages provided by SignBase, using the rcarbon package [@Crema_Bevan_2021]. The calibrated radiocarbon dates of the sampled data range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years. We assigned each object to a chronological phase following the date ranges recorded in @tbl-aur-table. The Transitional phase has `r trans_site_count` sites and `r trans_sign_count` sign types, the Proto-Aurignacian has `r proto_site_count` sites and `r proto_sign_count` sign types, the Early Aurignacian has `r early_site_count` sites and `r early_sign_count` sign types, and the Evolved Aurignacian has `r evolved_site_count` sites and `r evolved_sign_count` sign types. A Mantel test to evaluate the correlation between distance in time and distribution of sign types found a statistically significant correlation (Mantel statistic R = `r chron_r`, p = `r chron_p`). This confirms that our assumption that a meaningful change in sign types occurs over time during the Aurignacian period.

@@

Seriation analysis is the arrangement of data into a linear order to reveal patterns [@Hahsler_Hornik_Buchta]. We used the Brower-Kile [-@Brower_Kile_1988] seriation algorithm from the seriation package [@seriation_pckge], in which a unidimensional sequence is generated by reordering the rows and columns to group the presences along the diagonal. To investigate connectivity among clusters of sites in the seriation solution, we used network analysis to represent and quantify the relationship between sites [@Mills_2017]. In our network analysis, each node is a site, and each edge is a measure of sign type distribution similarity. We input the Jaccard similarity matrix for each phase, computed the network statistics, and plotted the results with the Fruchterman-Reingold layout [@Fruchterman_Reingold_1991], a force-directed graph layout for uniform edge lengths. Network analysis was performed using the vegan [@Vegan2001], statnet [@statnet_package] and igraph [@igraph_package] packages.

@@

To evaluate the strength of the groupings found in the seriation and network analyses, we used modularity and perMANOVA tests. The modularity test is a community structure network detection method, and quantifies the randomness of the groupings by measuring the fraction of the edges in the network that fall within groups, minus the expected value if the edges were placed at random [@Newman_2006; @Newman_Girvan_2004]. The results are expressed in a score of 0 to 1, with 0 meaning no significant subgroups, and 1 meaning perfect groupings. PerMANOVA quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017]. The results are expressed by an R-statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, and a p-value, which quantifies the probability of obtaining the observed difference if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance). The modularity test and perMANOVA were performed using the vegan [@Vegan2001] and igraph [@igraph_package] packages.

@@

The isolation-by-distance framework proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019]. If cultural difference can be solely explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases. To test this, we determined the correlation between sign distribution and geography using Mantel tests, run with a geographic distance matrix and a Jaccard similarity matrix. The test seeks to determine the level of correspondence between the matrices through permutational evaluation of the null distribution [@Mantel_Smouse], (here 1000 permutations), producing a Mantel R statistic and a p-value. The R statistic falls between -1, meaning a strong negative correlation, and 1, meaning a strong positive correlation, while a value of 0 means no correlation. The p-value quantifies the probability of obtaining the observed correlation if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance). The Mantel tests were performed using the vegan package [@Vegan2001]

@@

For each of the phases, seriation identifies at least two groups, one with a restricted range of sign types, and one with a broader range. The notch and line signs are the most prevalent sign types across all time periods. In the Transitional phase, there is a small range of sign types. The restricted group consists of Fumane, Pod Hradem, and Geissenklösterle, two of which only have a notch, and one of which has a notch and circumnotch. The broader group consists of Labeko Koba, Hohle Fels, and El Castillo. In the Proto-Aurignacian we see an expansion in both the number of sites and range of signs, with no overlap in sites from the Transitional phase, except for Geissenklösterle, which is present in both phases. The restricted group consists of Abri Pataud, Gatzarria, Hohlenstein-Stadel, and Mladeč, three of which only have notches and lines, and one also has a dashline. The broader group consists of Geissenklösterle, Spy, Vogelherd, Grotte du Renne, La Ferrassie, and Les Cottés, where we see a more consistent pattern of a more diverse range of signs, compared to the broader group of the Transitional phase. In the Early Aurignacian, we see appearance of completely new sites with no overlap with the Proto-Aurignacian sites, as well as increased diversity of patterns within the groups. The restricted group consists of Grottes de Fonds-de-Forêt (radnotch and notch), Riparo Bombrini (notch), and Grotte de la Verpillière I (notch and line), The broader group consists of Solutré, Hohle Fels, Castanet, Cellier, Blanchard, and Trou al'Wesse, which contain a wide range of signs in a variety of combinations. Lastly, in the Evolved Aurignacian, we see slightly more consistent groups, again with new sites with no overlap with the Early Aurignacian sites, except for Hohle Fels, which also appears in the Transitional phase. Vogelherd also reappears, having appeared earlier in the Proto-Aurignacian.The restricted group consists of Les Rois, Gargas, La Viña, Sirgenstein Cave, and Vindija Cave, covering the line, notch, and and obline signs. The broader group consists of Trou Magrite, Bockstein-Törle, Vogelherd, and Hohle Fels. Overall, across the four phases, the number of sites in the restricted group tends to be less than or equal to the number of sites in the broader group, with the exception of the final Evolved phase.
